\documentclass[a4paper]{article}

\usepackage[latin1]{inputenc}
\usepackage[english]{babel}
\usepackage{mathtools}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[nounderscore]{syntax}
\usepackage{bussproofs}
\usepackage{fancyvrb}

\VerbatimFootnotes

\setlength{\grammarparsep}{0pt}
\renewcommand{\syntleft}{\normalfont\itshape}
\renewcommand{\syntright}{}

\title{A Compiler for SPL with Custom Data Types}
\author{Tom Sanders and Gerco van Heerdt}

\begin{document}

\maketitle

\section{Project Description}

Our language has been extended with custom data types, generic field getters and setters, and constructor matching.
We explain this first using examples, after which we present the updated formal basis of our language in subsequent sections.

\subsection{Custom Data Types}

Our custom data types generalize both enumerations and record types.
As a simple example, the boolean type could have been defined as follows:
\begin{verbatim}
data Bool = True | False;\end{verbatim}
This type is denoted \verb+\Bool/+, so we might define a variable as follows:\footnote{%
    For the sake of the example we pretend here that there is no \verb+Bool+ type, for otherwise the value \verb+True+ would be of type \verb+Bool+ rather than \verb+\Bool/+, and a type error would occur.
}
\begin{verbatim}
\Bool/ a = True;\end{verbatim}
The constructors of custom types may have arguments:
\begin{verbatim}
data MaybeInt = Nothing | Just(Int val);\end{verbatim}
Moreover, type arguments may be declared, as we see in the following definitions of familiar types, taken from our standard library:
\begin{verbatim}
data List a = Nil | Cons(a hd, [a] tl);
data Tuple a b = Tuple(a fst, b snd);\end{verbatim}
Hence we recover generic lists and tuples as special cases of our construction.
A parameterized constructor behaves like a function from its arguments to the custom type, so A list with a single element $2$ may be defined as follows:
\begin{verbatim}
\List Int/ l = Cons(2, Nil);\end{verbatim}
By naming the constructor parameters (which is mandatory), we automatically unlock field getters and setters:
\begin{verbatim}
Int a = l.hd; // a = 2
l.tl = Cons(3, Nil); // l = Cons(2, Cons(3, Nil))\end{verbatim}

Note that in our language the types \verb+[a]+ and \verb+(a, b)+ are really synonyms for \verb+\List a/+ and \verb+\Tuple a b/+, respectively.
Furthermore, \verb+[]+ is actually \verb+Nil+, \verb+a : l+ translates to \verb+Cons(a, l)+, and \verb+(a, b)+ is equivalent to \verb+Tuple(a, b)+.

\subsection{Constructor Matching}

Additionally, we provide a construction to distinguish on the constructor value of an actual expression, which is much like a generalized if-statement.
The following code will print the number $2$:
\begin{verbatim}
\List Int/ l = Cons(2, Nil);
case l {
    Nil { print('?'); }
    Cons { print(l.hd); }
}\end{verbatim}
Note that there is no pattern matching on the constructor arguments available, but in most cases use of the field accessors is convenient enough (and sometimes even more convenient, if there are many constructor parameters).

We can now also implement the function \verb+isEmpty+ in the standard library:
\begin{verbatim}
Bool isEmpty([a] l) {
    case l {
        Nil { return True; }
        Cons { return False; }
    }
}\end{verbatim}

\subsection{Further Changes}

To avoid ambiguity, we need to make some additional restrictions.
Names of custom types and constructors must begin with an uppercase character; we have restricted variable and function names to begin with a lowercase character.

There is a minor other change: a variable declaration does not require an initial assignment.
This is not related to the project described above, but allows for nicer example code.
To avoid ambiguity with return statements (we do not use a tokenizer), polymorphic type identifiers may now only be single lowercase characters.

\section{Syntax}

Our new grammar is shown below.
\setlength{\grammarindent}{7.6em}
\begin{grammar}
    <SPL> ::= <Decl>+

    <Decl> ::= <DataDecl> | <VarDecl> | <FunDecl>

    <DataDecl> ::= `data' <uid> <lid>* `=' [\,<Constructor> (\,`|' <Constructor>\,)*\,] `;'

    <Constructor> ::= <uid> [\,`(' [\,<FArgs>\,] `)'\,]

    <VarDecl> ::= <Type> <lid> [\,`=' <Exp>\,] `;'

    <FunDecl> ::= <RetType> <lid> `(' [\,<FArgs>\,] `)' `{' <VarDecl>* <Stmt>* `}'

    <RetType> ::= <Type> | `Void'

    <Type> ::= <lower> | <BasicType> | `\\' <uid> <Type>* `/'
    \alt `(' <Type> `,' <Type> `)' | `[' <Type> `]'

    <BasicType> ::= `Int' | `Bool' | `Char'

    <FArgs> ::= <Type> <lid> [\,`,' <FArgs>\,]

    <Stmt> ::= `{' <Stmt>* `}' | <StmtId> `;' | `return' [\,<Exp>\,] `;'
    \alt `if' `(' <Exp> `)' <Stmt> [\,`else' <Stmt>\,]
    \alt `while' `(' <Exp> `)' <Stmt>
    \alt `case' <Exp> `{' (\,<uid> <Stmt>\,)* `}'

    <StmtId> ::= <lid> (\,[\,<Field>\,] `=' <Exp> | `(' [\,<ActArgs>\,] `)'\,)

    <Exp> ::= <Exp1> [\,`:' <Exp>\,]

    <Exp1> ::= <Exp2> (\,(\,`&&' | `||'\,) <Exp2>\,)*

    <Exp2> ::= <Exp3> (\,(\,`==' | `!='\,) <Exp3>\,)*

    <Exp3> ::= <Exp4> (\,(\,`<' | `>' | `<=' | `>='\,) <Exp4>\,)*

    <Exp4> ::= <Exp5> (\,(\,`+' | `-'\,) <Exp5>\,)*

    <Exp5> ::= <Exp6> (\,(\,`*' | `/' | `%'\,) <Exp6>\,)*

    <Exp6> ::= (\,`!' | `-'\,) <Exp6> | <NonOpExp>

    <NonOpExp> ::= <int> | <char> | <bool> | `[]' | <IdExp> | <ConsExp>
    \alt `(' <Exp> `)' | `(' <Exp> `,' <Exp> `)'

    <ConsExp> ::= <uid> [\,`(' [\,<ActArgs>\,] `)'\,]

    <IdExp> ::= <lid> (\,[\,<Field>\,] | `(' [\,<ActArgs>\,] `)'\,)

    <Field> ::= `.' <id> [\,<Field>\,]

    <ActArgs> ::= <Exp> [\,`,' <ActArgs>\,]

    <int> ::= [\,`-'\,] <digit>+

    <char> ::= `\'' <any> `\''

    <bool> ::= `False' | `True'

    <id> ::= <alpha> (\,`_' | <alphaNum>\,)*

    <lid> ::= <lower> (\,`_' | <alphaNum>\,)*

    <uid> ::= <upper> (\,`_' | <alphaNum>\,)*
\end{grammar}

Names of custom types and constructors must begin with an uppercase character; we have restricted variable and function names to begin with a lowercase character, and polymorphic type identifiers may only be single lowercase characters.
Note that there is a minor other change: a variable declaration does not require an initial assignment.

\section{Semantic Analysis}

Functions may be overloaded, but it is not allowed that multiple versions of a function apply for any possible way the function can be called.
Note that our implementation does not allow overloading the return type of a function at all.

Our newly added constructs induce some further checks.
It is not allowed to use the same field name twice within the same type definition, nor can the same type or constructor be declared twice.
In a case expression, the constructors for which a body is provided must correspond to constructors for the type of the expression that is evaluated.
A case body may be provided only once for every constructor in a case expression.
Finally, the type of the expression that is evaluated must be a custom type.

Internally, the following grammar is used to describe types:
\setlength{\grammarindent}{4.3em}
\begin{grammar}
    <Type> ::= Int | Bool | Char | Custom <uid> <Type>* | Poly <lower>
\end{grammar}

Before we can do any type checking, we need to define what a context is and how it is determined throughout the program.
For convenience, we use the notation $\Gamma = (\Gamma_v, \Gamma_f)$ expressing that a context $\Gamma$ is made up of a variable context $\Gamma_v$ and a function context $\Gamma_f$.
Functions and variables are added to the context in the usual way: variables defined in a subscope override ones defined on a higher level, and it is an error to redefine a variable in the same scope.
Note that both function arguments and function bodies introduce a new scope.
The set $\Gamma_v$ consists of pairs $(i, \sigma)$, where $i$ is a variable name and $\sigma$ its associated type; the set $\Gamma_f$ contains tuples $(i, \sigma \to \tau)$, where $i$ is a function name, $\sigma$ is an $n$-tuple containing the $n$ types of the function inputs, and $\tau$ is the output type.
For field getters and constructors we exploit the fact that normal function names always start with a lowercase character: encountering a constructor $\text{C}(\sigma_1\;f_1, \sigma_2\;f_2, \ldots, \sigma_n\;f_n)$ in a definition of a custom type T parameterized by a list of type variables $\mathit{as}$, we extend $\Gamma_f$ with the pairs $(@\text{C}, (\sigma_1, \sigma_2, \ldots, \sigma_n) \to \text{Custom T }\mathit{as}')$ and $(.f_i, \text{Custom T }\mathit{as}' \to \sigma_i)$ for all $1 \le i \le n$, where $\mathit{as}'$ is obtained from $\mathit{as}$ by interpreting every string as a polymorphic type.
Here and hereafter we use the denote by $.s$ and $@s$ the string resulting from prepending the character `.' or `@' to the string $s$.

Typechecking a function consists in ensuring that the returned values for all possible executions of the function (assuming just that it terminates) are compatible with the type given to it.
This has been updated to check case statements properly: for all possible constructors a value of the right type must be returned in the corresponding case body, unless subsequent code returns properly.
An execution not hitting a return statement is equivalent to it returning a variable of type Void in the end.
Of course there are no such variables, but this is in turn modeled by the return statement without argument.

\subsection{Typing rules}

We will type a simplified syntax, which is the one used internally to store a program rather than the one used for parsing.
The grammar describing our stored expressions is given below.
\setlength{\grammarindent}{4.1em}
\begin{grammar}
    <Exp> ::= Int $i$ | Bool $b$ | Char $c$ | $\ominus$<Exp> | <Exp> $\otimes$ <Exp>
    \alt Var <id> <Field>* | Fun <id> <Exp>* | Cons <id> <Exp>*
\end{grammar}
Here $\ominus$ stands for a unary operator, and $\otimes$ can be any binary operator; the value $i$ is an actual integer value, $b$ is any Boolean, and $c$ is a character.
Note that \synt{id} and \synt{Field} here are actually just arbitrary strings.

\begin{center}
    \AxiomC{}
    \UnaryInfC{$\Gamma \vdash \text{Int } i : \text{Int}$}
    \DisplayProof
    \qquad
    \AxiomC{}
    \UnaryInfC{$\Gamma \vdash \text{Bool } b : \text{Bool}$}
    \DisplayProof
    \qquad
    \AxiomC{}
    \UnaryInfC{$\Gamma \vdash \text{Char } c : \text{Char}$}
    \DisplayProof
\end{center}

\begin{center}
    \AxiomC{$\Gamma \vdash \text{Fun } @i\;e : \sigma$}
    \UnaryInfC{$\Gamma \vdash \text{Cons } i\;e : \sigma$}
    \DisplayProof
    \qquad
    \AxiomC{$(i, \sigma) \in \Gamma_v$}
    \UnaryInfC{$\Gamma \vdash \text{Var } i\;[\,] : \sigma$}
    \DisplayProof
\end{center}

\begin{center}
    \AxiomC{$(i, \sigma) \in \Gamma_v$}
    \AxiomC{$(.f, \sigma \to \tau) \in \Gamma_f$}
    \AxiomC{$(\Gamma_{v, i \mapsto \tau}, \Gamma_f) \vdash \text{Var } i\;r : \chi$}
    \TrinaryInfC{$\Gamma \vdash \text{Var } i\;(f :: r) : \chi$}
    \DisplayProof
\end{center}
Here $\Gamma_{v, i \mapsto \tau} = \{(j, t) \in \Gamma_v \mid j \neq i\} \cup \{(i, \tau)\}$.

\begin{center}
    \AxiomC{$\forall_{1 \le j \le n}\,\Gamma \vdash e_j : \sigma_j$}
    \AxiomC{$(i, (\sigma_1, \sigma_2, \ldots, \sigma_n) \to \tau) \in \Gamma_f$}
    \BinaryInfC{$\Gamma \vdash \text{Fun } i\;(e_1, e_2, \ldots, e_n) : \tau$}
    \DisplayProof
\end{center}

Next we have to specify all operators.

\begin{center}
    \AxiomC{$\Gamma \vdash e : \text{Int}$}
    \UnaryInfC{$\Gamma \vdash -e : \text{Int}$}
    \DisplayProof
    \qquad
    \AxiomC{$\Gamma \vdash e : \text{Bool}$}
    \UnaryInfC{$\Gamma \vdash\;!e : \text{Bool}$}
    \DisplayProof
    \qquad
    \AxiomC{$\Gamma \vdash e_1 : \sigma$}
    \AxiomC{$\Gamma \vdash e_2 : \sigma$}
    \BinaryInfC{$\Gamma \vdash e_1 \bullet e_2 : \sigma$}
    \DisplayProof
\end{center}
where $\bullet \in \{+, -, *, /, \%\}$.

\begin{center}
    \AxiomC{$\Gamma \vdash e : \sigma$}
    \AxiomC{$\Gamma \vdash l : [\sigma]$}
    \BinaryInfC{$\Gamma \vdash (e : l) : [\sigma]$}
    \DisplayProof
    \qquad
    \AxiomC{$\Gamma \vdash e_1 : \sigma$}
    \AxiomC{$\Gamma \vdash e_2 : \sigma$}
    \BinaryInfC{$\Gamma \vdash e_1 \diamond e_2 : \text{Bool}$}
    \DisplayProof
\end{center}
with $\diamond \in \{==, <, >, <=, >=, !\!=\}$.

\begin{center}
    \AxiomC{$\Gamma \vdash e_1 : \text{Bool}$}
    \AxiomC{$\Gamma \vdash e_2 : \text{Bool}$}
    \BinaryInfC{$\Gamma \vdash e_1 \text{ \&\& } e_2 : \text{Bool}$}
    \DisplayProof
    \qquad
    \AxiomC{$\Gamma \vdash e_1 : \text{Bool}$}
    \AxiomC{$\Gamma \vdash e_2 : \text{Bool}$}
    \BinaryInfC{$\Gamma \vdash e_1 \text{ \textbar\textbar{} } e_2 : \text{Bool}$}
    \DisplayProof
\end{center}

Finally, we have the following built-in functions, typed through context initialization:
\[
    \begin{array}{lcl}
        \mathit{print} & : & \sigma \to \text{Void} \\
        \mathit{read}  & : & \text{Int}
    \end{array}
\]

Note that we do not allow overloading or polymorphism purely on the return type, and so we choose to fix the type of the $\mathit{read}$ function to Int.
The reason for this is to avoid ambiguity: instead of as an expression a function call may appear as a statement, in which case the desired return type is not defined.
We further reject overloading when two versions of a function potentially apply to a certain same combination of argument types.

\section{Compilation Schemes}

We employ \emph{by-value} assignments and calling: a copy of an object is made on assignment.
For custom data types this value is a pointer to objects on the heap, but note that the heap is immutable because of the assignment convention.
Thus, an object passed to a function will never actually be modified by that function.

All objects are represented on the stack with a single value.
As mentioned above, values of custom data types are stored on the heap, by storing first the present arguments of the relevant constructor, followed by the constructor index in the current type.
% TODO: explain case distinctions

A program compiled by our compiler first assigns to the global variables in order their initial values, after which it calls the main function of which the return value, if any, is printed.
Our implementation uses a todo list maintaining the instances of functions that still need to be compiled.
Because of our design choices all types are known at compile time, so we can provide polymorphism simply through overloading as only a finite number of instances of a function can be called by a program.

On a more technical level, whenever we encounter a function call, we reserve room for all the parameters and variables in the relevant function and assign the values of the parameters.
The function is then called, and if it returns a value this will be stored in the return register.
For function calls as statements this can simply be ignored, but if the function call is used as an expression it will load the value of that register onto the stack (after the variables of the function are unlinked).

At the beginning of the program we copy the value of the stack pointer to another register (R7) such that we can calculate at any point the addresses of the global variables.

Many (cases of) built-in functions and operators are implemented through a standard library.
The remaining primitive cases (Int/Char printing, basic operators, etc.) are compiled inline.

\section{Build Instructions}

Building the project requires a Haskell compiler and a Make-like system.
Running \verb|make| in the project root will build the compiler in the \verb|build| folder.
The directory \verb|tests| contains all our tests, including ones for the error scenarios described earlier.
Programs can either be run manually by feeding SPL source files into the compiler with \verb|build/compiler < program.spl| or automatically by calling \verb|make tests| in the project root to run a set of predefined tests.
If no problems are encountered, the compiler generates the file \verb|a.ssm| containing the compiled program.

\end{document}
